---
title: "Analysis of class surveys"
subtitle: "If you want a subtitle put it here"
author: "Pratyush Rallapally, Anish Senthil, Satvik Talchuru, Andrew Guerra, Akhil Gorla"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. Overall, your report may be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
```

## Executive summary

Write a one-paragraph abstract summarizing what you did and your findings. It need not be comprehensive; try to highlight the most important or interesting outcomes.

## Data description

This dataset contains survey responses from UCSB undergraduate students enrolled in PSTAT197A, collected to capture their academic background, skills, and project interests. The survey collected personal information alongside self-reported proficiency and comfort in programming, mathematics, and statistics. Additional background measures include the number of upper-division courses completed, domain specialization, and prior research experience in data science and other areas.

The survey also recorded students’ project preferences, including preferred project type (industry vs. lab), programming language, domain, and data science areas of interest, as well as open-ended goals and comments for project placement. Several variables are binary (yes/no), ordinal, numeric, or categorical.

Overall, the dataset provides a detailed profile of students’ academic preparation, skill self-assessment, and project preferences, suitable for analyzing patterns in preparation, interests, and experiences among the PSTAT197A cohort. For our purposes, we believe the data provides a valuable snapshot of PSTAT majors, particularly those most motivated and engaged. While the survey cohort is naturally self-selected, this bias makes the data especially informative: it highlights where the department’s most engaged feel well-prepared or identify gaps in their skills. Understanding these patterns is crucial, as it allows the department to identify strengths and potential areas for curricular improvement.

## Questions of interest

Indicate the questions your analysis addresses. These should map one-to-one to your findings. Don't include questions you didn't answer or questions you started with and refined later. If you would rather frame them as goals or tasks rather than questions that is okay; just modify the header appropriately. However you frame what you've done, you may only have two or three items; that is fine. Provide an itemized or numbered list so that the reader can easily identify your objectives.

For example:

> *We sought to understand the relationship between coursework preparation, self-assessed technical abilities, and comfort level with technical skills. We also sought to understand whether the relationships differed depending on having had prior research experience. Our analysis addressed three questions:*
>
> 1.  *How should the coursework histories of students in the class be ordered from less preparation to extensive preparation?*
> 2.  *Is coursework preparation associated with increased self-assessed technical skill and comfort within the sample?*
> 3.  *Does prior research experience appear to be a substitute for coursework preparation with respect to self-assessed technical skill and comfort?*

Notice that the questions are precise but not overly technical.

3.  *To what extent do students’ self-assessed proficiency in programming, mathematics, and statistics align with their reported comfort in these domains?*

## Findings

Summarize your results. Don't try to explain every step you took; focus instead on providing the main data analytic outputs -- tables and figures -- and explaining clearly what they show. Clarify any important decisions you made in obtaining them. You can display codes if you like but it is not necessary.

We analyzed the relationship between students’ self-assessed proficiency and comfort in programming, mathematics, and statistics using the survey data. Proficiency responses were converted from categorical to numeric values, and for each domain, we fitted a simple linear regression model predicting comfort from proficiency. The results are summarized below:

```{r}
(lm_results)
```

In general, across all three domains, higher self-rated proficiency is associated with higher reported comfort. This expected relationship is the strongest in statistics, at almost twice the extent compared to programming and mathematics, who had nearly identical moderate associations.

We further explored these relationships through other visualizations, including scatterplots and a correlation heatmap.

```{r}
plot_data <- bind_rows(
  df %>% select(prof = prog.prof.num, comf = prog.comf) %>% mutate(domain = "Programming"),
  df %>% select(prof = math.prof.num, comf = math.comf) %>% mutate(domain = "Mathematics"),
  df %>% select(prof = stat.prof.num, comf = stat.comf) %>% mutate(domain = "Statistics")
)

ggplot(plot_data, aes(x = prof, y = comf)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(~ domain, scales = "free") +
  labs(
    x = "Self-rated Proficiency (1=beg, 2=int, 3=adv)",
    y = "Comfort Level (1-5)",
    title = "Linear Regression of Comfort on Proficiency by Domain"
  ) +
  theme_minimal()
```

In addition to seeing how the original data is clustered, the plots visually reaffirm our conclusions; within the same domain, students generally feel more confident the more proficient they are.

Thus far, the analysis has been domain-specific. To explore correlations across domains, we generated a correlation matrix, visualized in the form of a heatmap below:

```{r}
heatmap
```

From here, we see some interesting relations. Across the diagonal, we see the same earlier trend of higher proficiency = higher comfort when aligned in the same domain. However, cross-domain correlations are weaker, suggesting that proficiency in one area does not strongly predict comfort in another. Most notably, there appears to be no relationship whatsoever between math and programming. This pattern highlights that students’ perceived abilities are largely within the same domain, with comfort most closely tracking proficiency in the same area, particularly for statistics. The weaker cross-domain associations also suggest potential gaps in transferable confidence, indicating that even students who feel capable in one technical area may not feel equally comfortable applying those skills elsewhere. The relatively weaker alignments highlight gaps where even motivated students may feel less confident relative to their proficiency, necessitating actionable insight for curriculum planning and targeted support.
